Template for indicator documentation

### 2023-09-20 NDVI_natopen.Rmd
Small changes in the code to make it run in the book

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(downloader)
library(data.table)
library(lubridate)
library(sf)
library(plyr)
library(stringr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(betareg)
library(StepBeta)
library(glmmTMB)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r, echo=FALSE}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

@@

#### NiN data and geometry for Norway and the five regions
We read the NiN data for this document from a cache at the NINA server...
```{r, echo = F}
# Add NiN data from cache
nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
```

...but it can also be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
```{r, echo = T, eval = F}

url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="wb") 
unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data")
st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
                 layer="naturtyper_nin_omr")
```
And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
```{r message=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

reg <- st_read("data/regions.shp")%>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

# change region names to something R-friendly
reg$region
reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")

# combine the Norway and the region maps
regnor <- st_intersection(reg,nor)
rm(reg)
rm(nor)
```

#### NDVI data
All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE). 
```{r, echo = F}
# downloading the unprocessed NDVI data from a data-cache et the NINA server
df.s <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
df.m <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
df.l <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
```
##### Sentinel-2 NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
The image collection contains Sentinel imagery from June, July and August 2015-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r, echo = T, eval = F}
# Import Sentinel NDVI Data
df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names=TRUE) %>%
  map_df(~fread(.))
```

@@

```{r, echo = T, eval = F}
# Import MODIS NDVI Data
# MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
df.m$mean<-df.m$mean/10000
```

##### Landsat NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
The image collection contains Landsat imagery from June, July and August 1984-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r, echo = T, eval = F}
# Import Landsat NDVI Data
# Set up conditional file paths
dir <- substr(getwd(), 1,2)

path <- ifelse(dir == "C:", 
               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")

pData <- ifelse(dir == "C:", 
                "P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat",
                "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat")

# Fread doesn't like the weird path to the server version of the P drive
# hence this horrendous work around
files=list.files(pData, pattern = "*.csv", full.names = TRUE)

df_list<-list()# initialise a list of dataframes
# read in a dataframe in each slot of the df_list
for (i in files){
  name <- gsub("-",".",i)
  name <- gsub(".csv","",name)  
  i <- paste(i,sep="")
  df_list[[i]]=assign(name,read.csv(i, header=TRUE))
}

df.l<-bind_rows(df_list, .id = "column_label")
```

### Data handling
#### NiN data
```{r, echo = T, eval = F}
# fixing variable- and ecosystem-names with special characters
colnames(nin)
colnames(nin)[c(3,8,17,26,31,33,34)] <- c("hovedoekosystem","kartleggingsaar","noyaktighet",
                                          "omraadenavn","uk_naertruet","uk_sentraloekosystemfunksjon",
                                          "uk_spesieltdaarligkartlagt")
unique(nin$hovedoekosystem)

nin <- nin %>% mutate(hovedoekosystem = recode(hovedoekosystem, 
                                               "Våtmark" = 'Vaatmark',
                                               "Semi-naturlig mark" = 'Semi_naturlig',
                                               "Naturlig åpne områder i lavlandet" = 'Naturlig-aapent',
                                               "Naturlig åpne områder under skoggrensa" = 'Naturlig_aapent')) %>%
  mutate(validGeo = st_is_valid(SHAPE))

@@

nrow(nin)
# 95469 polygons altogether
nrow(nin %>%
       filter(grepl(',', ninkartleggingsenheter))
)
# 21094 polygons have more than 1 ecosystem type (they are separated by commas in the ninkartleggingsenheter-variable)

nrow(nin %>%
       filter(!grepl(',', ninkartleggingsenheter))
)
# 74375 polygons should have only 1 ecosystem type

# there's no information on the proportion of ecosystem types in the polygons, so we got to omit all polygons with multiple ecosystem types :(
nin <- nin %>%
  filter(!grepl(',', ninkartleggingsenheter))

# fix the content in the ninkartleggingsenheter-variable
summary(as.factor(nin$ninkartleggingsenheter))
# get rid of the NA- in the beginning
nin <- nin %>% mutate(ninkartleggingsenheter = str_remove(ninkartleggingsenheter, 'NA_'))
# making a main ecosystem type variable
nin <- nin %>% mutate(hovedtype = substr(ninkartleggingsenheter, 1, 3),
                      hovedtype = str_remove(hovedtype, '-'))
# checking mapping unit against main ecosystem type
nin[,c("hovedoekosystem","hovedtype")]
summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Vaatmark"]))
summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Semi_naturlig"]))
summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Naturlig_aapne"]))
summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Skog"]))
summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Fjell"]))
# making a new variable for the overarching ecosystem types based on the main ecosystem types
nin$hovedoekosystem.orig <- nin$hovedoekosystem

nin <- nin %>%
  mutate(hovedoekosystem = case_when(hovedtype %in% paste("V",c(1,3:7,9:10),sep="") ~ 'Vaatmark',
                                     hovedtype %in% paste("V",11:13,sep="") ~ 'Vaatmark_sterkt_endret',
                                     hovedtype %in% paste("T",c(31:34,40:41),sep="") ~ 'Semi_naturlig',
                                     hovedtype %in% paste("T",c(2,12,18,20:21),sep="") ~ 'Naturlig_aapent',
                                     hovedtype %in% c(paste("T",c(4,30,38),sep=""),paste("V",c(2,8),sep="")) ~ 'Skog',
                                     hovedtype %in% c(paste("T",c(3,7,9,10,14,22,26),sep=""),paste("V",c(6,7),sep="")) ~ 'Fjell',
                                     TRUE ~ 'NA'))

@@

nin <- nin %>% mutate(tilstand = recode(tilstand,
                                        "Dårlig" = "Redusert",
                                        "Svært redusert" = "Svaert_redusert"))
nin <- nin %>% mutate(tilstand_num = recode(tilstand,
                                        "God" = 0, "Moderat" = 1, 
                                        "Redusert" = 2, "Svaert_redusert" = 3))

@@

## filter out only natopen data
nin.natopen <- nin %>% 
  filter(hovedoekosystem %in% c('Naturlig_aapent')) %>%
  mutate(id = identifikasjon_lokalid) %>%
  filter(validGeo) %>%
  drop_na(tilstand) %>%
  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, tilstand_num, ninbeskrivelsesvariabler, kartleggingsaar)

# merge NiN-data with region
nin.natopen = st_join(nin.natopen, regnor, left = TRUE)
nin.natopen

colnames(nin.natopen)[c(1,10)] <- c("id","region_id")

@@

nin.natopen <- nin.natopen %>% 
  mutate(area_meters_nin = st_area(nin.natopen)
  )
# check and edit the order of regions
levels(nin.natopen$region)
nin.natopen$region <- as.factor(nin.natopen$region)
levels(nin.natopen$region)
nin.natopen$region <- factor(nin.natopen$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
levels(nin.natopen$region)

# splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
nin.natopen2 <- nin.natopen %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
  separate(col = ninbeskrivelsesvariabler,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  mutate(NiN_variable_code = as.factor(NiN_variable_code)) %>%
  subset(select = -ninbeskrivelsesvariabler)
nin.natopen2 <- as.data.frame(nin.natopen2)

```

```{r, echo = F}

#load processed NiN-data from cache
nin.natopen <- readRDS(paste0(here::here(),"/data/cache/nin.natopen.RDS"))
nin.natopen2 <- readRDS(paste0(here::here(),"/data/cache/nin.natopen2.RDS"))

```

#### NDVI data
```{r, results='hide', message=FALSE}
## Sentinel-2
# join nin.natopen & Sentinel NDVI data
SentinelNDVI.natopen <- full_join(nin.natopen, df.s, by="id")
#summary(SentinelNDVI.natopen)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(SentinelNDVI.natopen))
#summary(SentinelNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(hovedtype))
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(mean))
#summary(SentinelNDVI.natopen)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.natopen)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.natopen)
# split date into year, month & day
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# add column for sub-ecosystem types
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
                                                        subtype = str_remove(subtype, '-'),
                                                        subtype = str_remove(subtype, '-'))
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

@@

## MODIS
# join nin.natopen and Modis NDVI data
ModisNDVI.natopen <- full_join(nin.natopen, df.m, by="id")

#summary(ModisNDVI.natopen)
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(ModisNDVI.natopen))
#summary(ModisNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(hovedtype))
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(mean))
#summary(ModisNDVI.natopen)
# get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
dim(ModisNDVI.natopen)
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(as.numeric(area_meters) >= 62500)
dim(ModisNDVI.natopen)
# split date into year, month & day
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# we are using max NDVI per year in every NiN polygon
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))
#summary(ModisNDVI.natopen)

## Landsat
# join nin.natopen & Landsat NDVI data
LandsatNDVI.natopen <- full_join(nin.natopen, df.l, by="id")

#summary(LandsatNDVI.natopen)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  mutate(hovedoekosystem = as.factor(hovedoekosystem),
         hovedtype = as.factor(hovedtype),
         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
         lokalitetskvalitet = as.factor(lokalitetskvalitet),
         tilstand = as.factor(tilstand),
         area_meters = st_area(LandsatNDVI.natopen))
#summary(LandsatNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(hovedtype))
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(mean))
#summary(LandsatNDVI.natopen)
# get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
dim(LandsatNDVI.natopen)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(as.numeric(area_meters) >= 900)
dim(LandsatNDVI.natopen)
# split date into year, month & day
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  dplyr::mutate(year = lubridate::year(date), 
                month = lubridate::month(date), 
                day = lubridate::day(date))
# we are using max NDVI per year in every NiN polygon
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm=TRUE))

#summary(LandsatNDVI.natopen)
```

>>>>

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(downloader)
library(data.table)
library(lubridate)
library(sf)
library(plyr)
library(stringr)
library(tidyverse)
library(RColorBrewer)
library(gridExtra)
library(ggridges)
library(tmap)
library(tmaptools)
library(betareg)
library(StepBeta)
library(glmmTMB)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r, echo=FALSE}
metaData <- data.frame(Ecosystem,
  "Økologisk egenskap" = Egenskap,
  "ECT class" = ECT
)
knitr::kable(metaData)
```

@@

#### NiN data and geometry for Norway and the five regions
We read the NiN data for this document from a cache at the NINA server...
```{r, echo = F}
# Add NiN data from cache
nin <- readRDS(paste0(here::here(), "/data/cache/nin.RDS"))
```

...but it can also be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
```{r, echo = T, eval = F}
url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
download(url, dest = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode = "wb")
unzip("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip",
  exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data"
)
st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
  layer = "naturtyper_nin_omr"
)
```

And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
```{r message=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

reg <- st_read("data/regions.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

# change region names to something R-friendly
reg$region
reg$region <- c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway")

# combine the Norway and the region maps
regnor <- st_intersection(reg, nor)
rm(reg)
rm(nor)
```

#### NDVI data
All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE). 
```{r, echo = F}
# downloading the unprocessed NDVI data from a data-cache et the NINA server
df.s <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
df.m <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
df.l <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
```
##### Sentinel-2 NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
The image collection contains Sentinel imagery from June, July and August 2015-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r, echo = T, eval = F}
# Import Sentinel NDVI Data
df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names = TRUE) %>%
  map_df(~ fread(.))
```

@@

```{r, echo = T, eval = F}
# Import MODIS NDVI Data
# MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
df.m <- read.csv("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_MODIS/modis_ndvi_ts_2000_2022.csv")
df.m$mean <- df.m$mean / 10000
```

##### Landsat NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
The image collection contains Landsat imagery from June, July and August 1984-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r, echo = T, eval = F}
# Import Landsat NDVI Data
# Set up conditional file paths
dir <- substr(getwd(), 1, 2)

path <- ifelse(dir == "C:",
  "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
  "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
)

pData <- ifelse(dir == "C:",
  "P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat",
  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat"
)

# Fread doesn't like the weird path to the server version of the P drive
# hence this horrendous work around
files <- list.files(pData, pattern = "*.csv", full.names = TRUE)

df_list <- list() # initialise a list of dataframes
# read in a dataframe in each slot of the df_list
for (i in files) {
  name <- gsub("-", ".", i)
  name <- gsub(".csv", "", name)
  i <- paste(i, sep = "")
  df_list[[i]] <- assign(name, read.csv(i, header = TRUE))
}

df.l <- bind_rows(df_list, .id = "column_label")
```

### Data handling
#### NiN data
```{r, echo = T, eval = F}
# fixing variable- and ecosystem-names with special characters
colnames(nin)
colnames(nin)[c(3, 8, 17, 26, 31, 33, 34)] <- c(
  "hovedoekosystem", "kartleggingsaar", "noyaktighet",
  "omraadenavn", "uk_naertruet", "uk_sentraloekosystemfunksjon",
  "uk_spesieltdaarligkartlagt"
)
unique(nin$hovedoekosystem)

nin <- nin %>%
  mutate(hovedoekosystem = recode(hovedoekosystem,
    "Våtmark" = "Vaatmark",
    "Semi-naturlig mark" = "Semi_naturlig",
    "Naturlig åpne områder i lavlandet" = "Naturlig-aapent",
    "Naturlig åpne områder under skoggrensa" = "Naturlig_aapent"
  )) %>%
  mutate(validGeo = st_is_valid(SHAPE))

@@

nrow(nin)
# 95469 polygons altogether
nrow(nin %>%
  filter(grepl(",", ninkartleggingsenheter)))
# 21094 polygons have more than 1 ecosystem type (they are separated by commas in the ninkartleggingsenheter-variable)

nrow(nin %>%
  filter(!grepl(",", ninkartleggingsenheter)))
# 74375 polygons should have only 1 ecosystem type

# there's no information on the proportion of ecosystem types in the polygons, so we got to omit all polygons with multiple ecosystem types :(
nin <- nin %>%
  filter(!grepl(",", ninkartleggingsenheter))

# fix the content in the ninkartleggingsenheter-variable
summary(as.factor(nin$ninkartleggingsenheter))
# get rid of the NA- in the beginning
nin <- nin %>% mutate(ninkartleggingsenheter = str_remove(ninkartleggingsenheter, "NA_"))
# making a main ecosystem type variable
nin <- nin %>% mutate(
  hovedtype = substr(ninkartleggingsenheter, 1, 3),
  hovedtype = str_remove(hovedtype, "-")
)
# checking mapping unit against main ecosystem type
nin[, c("hovedoekosystem", "hovedtype")]
summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Vaatmark"]))
summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Semi_naturlig"]))
summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Naturlig_aapne"]))
summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Skog"]))
summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Fjell"]))
# making a new variable for the overarching ecosystem types based on the main ecosystem types
nin$hovedoekosystem.orig <- nin$hovedoekosystem

nin <- nin %>%
  mutate(hovedoekosystem = case_when(
    hovedtype %in% paste("V", c(1, 3:7, 9:10), sep = "") ~ "Vaatmark",
    hovedtype %in% paste("V", 11:13, sep = "") ~ "Vaatmark_sterkt_endret",
    hovedtype %in% paste("T", c(31:34, 40:41), sep = "") ~ "Semi_naturlig",
    hovedtype %in% paste("T", c(2, 12, 18, 20:21), sep = "") ~ "Naturlig_aapent",
    hovedtype %in% c(paste("T", c(4, 30, 38), sep = ""), paste("V", c(2, 8), sep = "")) ~ "Skog",
    hovedtype %in% c(paste("T", c(3, 7, 9, 10, 14, 22, 26), sep = ""), paste("V", c(6, 7), sep = "")) ~ "Fjell",
    TRUE ~ "NA"
  ))

@@

nin <- nin %>% mutate(tilstand = recode(tilstand,
  "Dårlig" = "Redusert",
  "Svært redusert" = "Svaert_redusert"
))
nin <- nin %>% mutate(tilstand_num = recode(tilstand,
  "God" = 0, "Moderat" = 1,
  "Redusert" = 2, "Svaert_redusert" = 3
))

@@

## filter out only natopen data
nin.natopen <- nin %>%
  filter(hovedoekosystem %in% c("Naturlig_aapent")) %>%
  mutate(id = identifikasjon_lokalid) %>%
  filter(validGeo) %>%
  drop_na(tilstand) %>%
  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, tilstand_num, ninbeskrivelsesvariabler, kartleggingsaar)

# merge NiN-data with region
nin.natopen <- st_join(nin.natopen, regnor, left = TRUE)
nin.natopen

colnames(nin.natopen)[c(1, 10)] <- c("id", "region_id")

@@

nin.natopen <- nin.natopen %>%
  mutate(area_meters_nin = st_area(nin.natopen))
# check and edit the order of regions
levels(nin.natopen$region)
nin.natopen$region <- as.factor(nin.natopen$region)
levels(nin.natopen$region)
nin.natopen$region <- factor(nin.natopen$region, levels = c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway"))
levels(nin.natopen$region)

# splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
nin.natopen2 <- nin.natopen %>%
  separate_rows(ninbeskrivelsesvariabler, sep = ",") %>%
  separate(
    col = ninbeskrivelsesvariabler,
    into = c("NiN_variable_code", "NiN_variable_value"),
    sep = "_",
    remove = F
  ) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  mutate(NiN_variable_code = as.factor(NiN_variable_code)) %>%
  subset(select = -ninbeskrivelsesvariabler)
nin.natopen2 <- as.data.frame(nin.natopen2)
```

```{r, echo = F}
# load processed NiN-data from cache
nin.natopen <- readRDS(paste0(here::here(), "/data/cache/nin.natopen.RDS"))
nin.natopen2 <- readRDS(paste0(here::here(), "/data/cache/nin.natopen2.RDS"))
```

#### NDVI data
```{r, results='hide', message=FALSE}
## Sentinel-2
# join nin.natopen & Sentinel NDVI data
SentinelNDVI.natopen <- full_join(nin.natopen, df.s, by = "id")
# summary(SentinelNDVI.natopen)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(SentinelNDVI.natopen)
  )
# summary(SentinelNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(hovedtype))
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(mean))
# summary(SentinelNDVI.natopen)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.natopen)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.natopen)
# split date into year, month & day
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# add column for sub-ecosystem types
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% mutate(
  subtype = substring(ninkartleggingsenheter, 4),
  subtype = str_remove(subtype, "-"),
  subtype = str_remove(subtype, "-")
)
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

@@

## MODIS
# join nin.natopen and Modis NDVI data
ModisNDVI.natopen <- full_join(nin.natopen, df.m, by = "id")

# summary(ModisNDVI.natopen)
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(ModisNDVI.natopen)
  )
# summary(ModisNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(hovedtype))
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(mean))
# summary(ModisNDVI.natopen)
# get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
dim(ModisNDVI.natopen)
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(as.numeric(area_meters) >= 62500)
dim(ModisNDVI.natopen)
# split date into year, month & day
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# we are using max NDVI per year in every NiN polygon
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))
# summary(ModisNDVI.natopen)

## Landsat
# join nin.natopen & Landsat NDVI data
LandsatNDVI.natopen <- full_join(nin.natopen, df.l, by = "id")

# summary(LandsatNDVI.natopen)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(LandsatNDVI.natopen)
  )
# summary(LandsatNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(hovedtype))
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(mean))
# summary(LandsatNDVI.natopen)
# get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
dim(LandsatNDVI.natopen)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(as.numeric(area_meters) >= 900)
dim(LandsatNDVI.natopen)
# split date into year, month & day
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# we are using max NDVI per year in every NiN polygon
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

# summary(LandsatNDVI.natopen)
```
